---
title: "Health & Income Research Analysis"
---

```{r}
#| label: setup
#| include: false
#| 
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
load(here("data","data_constructed","analytical_data.RData))
```


```{r}
## Steps missed during cleaning
cleaned_dataset
cleaned_dataset <- cleaned_dataset |>
  filter(!is.na(HasCondition))
cleaned_dataset # eliminate entries that did not answer if they have condition
cleaned_dataset <- cleaned_dataset |>
  mutate(
    ConditionGroup = case_when(
      HasCondition == 0 ~ "No Condition",
      HasCondition == 1 ~ "Has Condition",
      TRUE ~ NA_character_
    )
  )
cleaned_dataset <- cleaned_dataset |> # eliminate invalid income values
    filter(!Income %in% c(-1, -2, -9))
```

```{r}
## Baseline Income Distributions
#| label: fig-histogram 1
#| fig-cap: Histogram of Income Distribution across all Participants

ggplot(cleaned_dataset, aes(x = Income)) +
  geom_histogram(binwidth = 250) +
  labs(title = "General Income Distribution", x = "Income ($)", y = "Count")

#| label: fig-histogram 2
#| fig-cap: Histogram of Income Distribution with No Health Condition
  
  no_condition <- cleaned_dataset |>
  filter(HasCondition == 0)

  ggplot(no_condition, aes(x = Income)) +
  geom_histogram(binwidth = 250) +
  labs(title = "Has Condition Income Distribution", x = "Income ($)", y = "Count")

#| label: fig-histogram 3
#| fig-cap: Histogram of Income Distribution with Health Condition
  has_condition <- cleaned_dataset |>
  filter(HasCondition == 1)

  ggplot(has_condition, aes(x = Income)) +
  geom_histogram(binwidth = 250) +
  labs(title = "No Condition Income Distribution", x = "Income ($)", y = "Count")
  
ggplot(cleaned_dataset, aes(x = Income, fill = ConditionGroup)) +
  geom_histogram(position = "identity", alpha = 0.5, binwidth = 250) +
  labs(title = "Income Distributions by Health Condition",
       x = "Income ($)",
       y = "Count") +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  theme_minimal()

## Income Summary Comparison
income_summary <- cleaned_dataset |>
  group_by(ConditionGroup) |>
  summarize(
    count = n(),
    mean_income = mean(Income, na.rm = TRUE),
    median_income = median(Income, na.rm = TRUE),
    min_income = min(Income, na.rm = TRUE),
    max_income = max(Income, na.rm = TRUE),
    income_range = max_income - min_income
  )
income_summary
```
When looking at the income summaries, the most noticable difference in the 
range in income. The group with no health condition has a maximum value of 25000
while the group with only has 8000 (granted this 25000 figure is an outlier). 
Additionally the median and mean incomes are both lower for the group with a 
health condition. When looking visually, both distributions look pretty similar:
right skewed distributions that taper off as you approach higher income values. 

